---
title: ' TELCO CUSTOMER CHURN PREDICTION USING MACHINE LEARNING ALGORITHMS'
author: "Precious Ebu"
date: "6/22/2021"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#activating required libraries 
```{r}

library(tidyverse)
library(psych)
library(ggpubr)
library(ggplot2)
library(readxl)
library(stringr)
library(lattice)
library(caret)
library(randomForest)
library(Boruta)
library(car)
```

#Importing the dataset 
```{r}
telcom <- read_csv("C:\\Users\\user\\Documents\\DATALAB\\Datasets\\Telcom Dataset.csv")
```

#inspecting the dataset
```{r}
head(telcom)
view(telcom)
```

#getting more insight on the dataset
```{r}
str(telcom) #checking the structure of the dataset

glimpse(telcom) #getting a glimpse of the dataset

summary(telcom) #getting a summary of the dataset
```
#This dataset has 7043 observations and 21 variables. Also observed the presence of NAs, wrong datatypes, NULL values. 


#Checking unique values of each character variable
```{r}
unique(telcom$gender) 
```

```{r}
unique(telcom$SeniorCitizen)
```

```{r}
unique(telcom$Partner)  
```

```{r}
unique(telcom$Dependents)  
```

```{r}
unique(telcom$PhoneService)  
```

```{r}
unique(telcom$MultipleLines)  
```

```{r}
unique(telcom$InternetService)
```

```{r}
unique(telcom$OnlineSecurity) 
```

```{r}
unique(telcom$OnlineBackup) 
```

```{r}
unique(telcom$DeviceProtection) 
```

```{r}
unique(telcom$TechSupport) 
```

```{r}
unique(telcom$StreamingTV) 
```

```{r}
unique(telcom$StreamingMovies) 
```

```{r}
unique(telcom$Contract) 
```

```{r}
unique(telcom$PaperlessBilling) 
```

```{r}
unique(telcom$Churn)
```
#Observed some variables are categorical 

DATA TRANSFORMATION
#Converting character variables to factors 
```{r}

telcom$gender <- as.factor(telcom$gender)
telcom$SeniorCitizen <- as.factor(telcom$SeniorCitizen)
telcom$Partner <- as.factor(telcom$Partner)
telcom$Dependents <- as.factor(telcom$Dependents)
telcom$PhoneService <- as.factor(telcom$PhoneService)
telcom$MultipleLines <- as.factor(telcom$MultipleLines)
telcom$InternetService <- as.factor(telcom$InternetService)
telcom$OnlineSecurity <- as.factor(telcom$OnlineSecurity)
telcom$OnlineBackup <- as.factor(telcom$OnlineBackup)
telcom$DeviceProtection <- as.factor(telcom$DeviceProtection)
telcom$TechSupport <- as.factor(telcom$TechSupport)
telcom$StreamingTV <- as.factor(telcom$StreamingTV)
telcom$StreamingMovies <- as.factor(telcom$StreamingMovies)
telcom$Contract <- as.factor(telcom$Contract)
telcom$PaperlessBilling <- as.factor(telcom$PaperlessBilling)
telcom$PaymentMethod <- as.factor(telcom$PaymentMethod)
telcom$Churn <- as.factor(telcom$Churn)
```

#confirming the datatypes conversion 
```{r}
str(telcom) 
```

#converting numeric column "tenure" to integer
```{r}
telcom$tenure <- as.integer(telcom$tenure) 
```

#confirming the datatype conversion
```{r}
class(telcom$tenure) 
```

DATA CLEANING
#replacing duplicates and NULL values
```{r}
telcom$InternetService <-   str_replace(telcom$InternetService, 'dsl', 'DSL')

telcom$Contract <- str_replace(telcom$Contract, '2 years', 'Two year')

telcom$OnlineSecurity <- str_replace(telcom$OnlineSecurity, 'NO', 'No')

telcom$OnlineSecurity <- str_replace(telcom$OnlineSecurity, 'No Internet Service', 'No')

telcom$OnlineBackup <- str_replace(telcom$OnlineBackup, 'NO', 'No')

telcom$StreamingMovies <- str_replace(telcom$StreamingMovies, 'NO', 'No')

telcom$TechSupport <- str_replace(telcom$TechSupport, 'YES', 'Yes')

telcom$PaymentMethod <- str_replace(telcom$PaymentMethod, 'Electronic_check', 'Electronic check')

telcom[telcom == "\\"] <- NA

telcom[telcom == "No internet service"] <- "No"

telcom[,(15 == "NO")] <- "No"
```

#using the apply function to effect changes in desired columns
```{r}
apply(telcom[, 2:18], 2, function(x) unique(x))
```

#Checking for NAs (Missing values) in the dataset 
```{r}
which(is.na(telcom)) 
sum(is.na(telcom)) 
```

#Checking sum of NAs in each column
```{r}
colSums(is.na(telcom)) 
```

#Checking the descriptive statistics of the numeric columns
```{r}
mean(telcom$tenure, na.rm = TRUE) 

median(telcom$tenure, na.rm = TRUE) 

sd(telcom$tenure, na.rm = TRUE) 
```
#Mean and Median do not coincide indicating that the distribution is not normal
```{r}
mean(telcom$MonthlyCharges, na.rm = TRUE)  

median(telcom$MonthlyCharges, na.rm = TRUE) 

sd(telcom$MonthlyCharges, na.rm = TRUE) 
```
#Mean and Median do not coincide indicating that the distribution is not normal
```{r}
mean(telcom$TotalCharges, na.rm = TRUE) 

median(telcom$TotalCharges, na.rm = TRUE) 

sd(median(telcom$TotalCharges, na.rm = TRUE)) 
```
#Mean and Median do not coincide indicating that the distribution is not normal

REPLACING MISSING VALUES
#replacing NAs in tenure(numeric column) with the column median
```{r}
telcom$tenure[is.na(telcom$tenure)] <- median(telcom$tenure, na.rm = TRUE)
```

#confirming the replacement
```{r}
which(is.na(telcom$tenure)) 
```
#replacing NAs in MonthlyCharges(numeric column) with the column mean 
```{r}
telcom$MonthlyCharges[is.na(telcom$MonthlyCharges)] <- mean(telcom$MonthlyCharges, na.rm = TRUE)
```

#confirming the replacement
```{r}
which(is.na(telcom$MonthlyCharges)) 
```
#replacing the missing values in Totalcharges with a constant 
```{r}
telcom$TotalCharges[is.na(telcom$TotalCharges)] <- 2284
```

#confirming the replacement
```{r}
which(is.na(telcom$TotalCharges)) 
```

#Creating tables to check for modes of the character columns
```{r}
table(telcom$Partner)
table(telcom$Dependents)
table(telcom$MultipleLines)
table(telcom$InternetService)
table(telcom$DeviceProtection)
table(telcom$OnlineBackup)
table(telcom$DeviceProtection)
table(telcom$TechSupport)
table(telcom$StreamingTV)
table(telcom$StreamingMovies)
table(telcom$Contract)
table(telcom$PaperlessBilling)
table(telcom$PaymentMethod)

```

#replacing NAs in the character columns with their respective modes
```{r}
telcom$Partner[is.na(telcom$Partner)] <- "No"
telcom$Dependents[is.na(telcom$Dependents)] <- "No"
telcom$MultipleLines[is.na(telcom$MultipleLines)] <- "No"
telcom$InternetService[is.na(telcom$InternetService)] <- "Fiber optic"
telcom$DeviceProtection[is.na(telcom$DeviceProtection)] <- "No"
telcom$OnlineBackup[is.na(telcom$OnlineBackup)] <- "No"
telcom$DeviceProtection[is.na(telcom$DeviceProtection)] <- "No"
telcom$TechSupport[is.na(telcom$TechSupport)] <- "No"
telcom$Contract[is.na(telcom$Contract)] <- "Month-to-month"
telcom$StreamingTV[is.na(telcom$StreamingTV)] <- "No"
telcom$StreamingMovies[is.na(telcom$StreamingMovies)] <- "No"
telcom$PaperlessBilling[is.na(telcom$PaperlessBilling)] <- "Yes"
telcom$PaymentMethod[is.na(telcom$PaymentMethod)] <- "Electronic check"
```
#Confirming NAs have been replaced
```{r}
which(is.na(telcom)) 
colSums(is.na(telcom))
sum(is.na(telcom)) 
```

#Checking the desciptive statistics of the dataset after replacing missing values
```{r}
describe(telcom) 
```
#No significant change observed between the statistics before replacing NAs and after.


DETECTION OF OUTLIERS
#detecting the presence of outliers using box plots
```{r}
boxplot(telcom$tenure)
```
#2 outliers detected

```{r}
boxplot(telcom$MonthlyCharges)
```
#no outliers observed

```{r}
boxplot(telcom$TotalCharges)
```
#no outliers observed

#saving outliers into a variable
```{r}
outliers <- boxplot(telcom$tenure, plot = FALSE)$out 
print(outliers) 
```

#checking the position of outliers
```{r}
telcom[which(telcom$tenure %in% outliers),] 
```

#dealing with outliers
```{r}
tenure_mean <- round(mean(telcom$tenure))
tenure_mean
```
#replacing outliers in tenure with the mean
```{r}
telcom$tenure[c(16,4000)] <- tenure_mean 
```

#confirming outliers have been dealt with
```{r}
boxplot(telcom$tenure) 
```

#Data is now clean

FEATURE ENGINEERING
#engineering a new column from tenure and Monthly charges
```{r}
TotalChargesCal <- ((telcom$tenure)*(telcom$MonthlyCharges)) 
```

#Column binding and confirming the engineered column
```{r}
telcom$TotalChargesCal <- c(TotalChargesCal) 
view(telcom) 
```

#checking the datatype for the new feature
```{r}
class(telcom$TotalChargesCal)
```

#getting the descriptive statistics
```{r}
mean(telcom$TotalChargesCal, na.rm = TRUE) 

median(telcom$TotalChargesCal, na.rm = TRUE) 

sd(telcom$TotalChargesCal, na.rm = TRUE) 
```

EXPLORATORY DATA ANALYSIS

#Data visualization using scatter plot 
```{r}
plot(tenure ~ MonthlyCharges, data = telcom) 
```
#from the scatter plot, a cluster of values is observed showing a linear relationship between tenure and monthly charges causing both to increase simultaneously

#visualizing the target variable churn using bar plot
```{r}
ggplot(telcom, aes(x = Churn))+ 
  geom_bar(mapping = aes(fill=Churn)) + geom_text(aes(y=..count..-200, 
                label = paste0(round(prop.table(..count..),4) * 100, '%')), 
            stat = 'count', position = position_dodge(.1), size =3)
```
#this plot shows that only about one quarter of the sample population have churned.

#data visualization using bar charts of each independent variable against the target variable "Churn"
```{r}
variables <-list("Partner", "Dependents", "SeniorCitizen", "InternetService", "OnlineBackup", "OnlineSecurity", "DeviceProtection", "TechSupport", "StreamingMovies", "StreamingTV", "Contract", "PaperlessBilling", "PaymentMethod") 
plot <- list()

 for (i in variables){
 plot <-  ggplot(data = telcom, aes_string(x=i, fill = as.factor(telcom$Churn)))+
  geom_bar(position = "stack") + 
  geom_text(aes(y=..count..-200, 
                label = paste0(round(prop.table(..count..),4) * 100, '%')), 
            stat = 'count', position = position_dodge(.1), size =3)
   print(plot)
}
```

```{r}
ggplot(data = telcom, aes(x=tenure, fill = Churn)) + geom_histogram(binwidth = 1.5) +labs(x= "Months", title = "Churn by tenure rate")
```

```{r}
ggplot(data = telcom, aes(x=MonthlyCharges, fill = Churn)) + geom_histogram(binwidth = 1.5) +labs(x= "Charges", title = "Churn by Monthly Charges")
```

```{r}
ggplot(data = telcom, aes(x=TotalCharges, fill = Churn)) + geom_histogram(binwidth = 100) +labs(x= "Amount (binwidth=100)", title = "Churn by Totalcharges")
```

```{r}
cor(telcom$tenure, telcom$TotalChargesCal)
```

```{r}
cor(telcom$tenure, telcom$MonthlyCharges)
```
the correlation coefficient is 0.24, this shows weak positive correlation exists between tenure and Monthly Charges
```{r}
cor(telcom$tenure, telcom$TotalCharges)
```

```{r}
cor(telcom$MonthlyCharges, telcom$TotalCharges)
```
```{r}
cor(telcom$MonthlyCharges, telcom$TotalChargesCal)
```
```{r}
cor(telcom$TotalCharges, telcom$TotalChargesCal)
```

```{r}
chisq.test(telcom$gender, telcom$Churn)
```

```{r}
chisq.test(telcom$SeniorCitizen, telcom$Churn)
```

```{r}
chisq.test(telcom$Partner, telcom$Churn)
```

```{r}
chisq.test(telcom$Dependents, telcom$Churn)
```

```{r}
chisq.test(telcom$PhoneService, telcom$Churn)
```

```{r}
chisq.test(telcom$MultipleLines, telcom$Churn)
```

```{r}
chisq.test(telcom$InternetService, telcom$Churn)
```

```{r}
chisq.test(telcom$OnlineSecurity, telcom$Churn)
```

```{r}
chisq.test(telcom$OnlineBackup, telcom$Churn)
```

```{r}
chisq.test(telcom$DeviceProtection, telcom$Churn)
```

```{r}
chisq.test(telcom$TechSupport, telcom$Churn)
```

```{r}
chisq.test(telcom$StreamingTV, telcom$Churn)
```

```{r}
chisq.test(telcom$StreamingMovies, telcom$Churn)
```


```{r}
chisq.test(telcom$Contract, telcom$Churn)
```


```{r}
chisq.test(telcom$PaperlessBilling, telcom$Churn)
```


```{r}
chisq.test(telcom$PaymentMethod, telcom$Churn)
```


```{r}
result <- aov(MonthlyCharges ~ Churn, data = telcom)
summary(result)
```


```{r}
LogRegChurn <- glm(Churn~MonthlyCharges, data = telcom, family = binomial(link = "logit"))
```

```{r}
summary(LogRegChurn)
```


```{r}
telcom1 <- data.frame(telcom[,c(-1,-20,-22)]) 
view(telcom1)
```

#Splitting the dataset into training and testing data
```{r}
Churn_yes <- telcom1[which(telcom1$Churn=="Yes"),]
Churn_no <- telcom1[which(telcom1$Churn=="No"),]
```

#creating Training data from 75% of the dataset and balancing my yes and no values
```{r}
set.seed(1500)
train1 <- sample(1:nrow(Churn_yes), 0.75*nrow(Churn_yes))
train2 <- sample(1:nrow(Churn_no), 0.75*nrow(Churn_no))
```


```{r}
train_y <- Churn_yes[train1,]
view(train_y)
train_n <- Churn_no[train2,]
view(train_n)
```

#row binding both partitions
```{r}
train_data <- rbind(train_y,train_n)
```

#shuffling the training data
```{r}
rows <- sample(nrow(train_data))
train_data <- train_data[rows, ]
```

#viewing the split training data
```{r}
view(train_data) 
```

#creating the test data
```{r}
test_y <- Churn_yes[-train1,] #contains 25% of churn "yes"
test_n <- Churn_no[-train2,] #contains 25% of churn "no"
test_data <- rbind(test_y, test_n)
```

#Shuffling the test data
```{r}
rows <- sample(nrow(test_data))
test_data <- test_data[rows, ]
view(test_data)
```

#Viewing the split test data
```{r}
view(test_data)
```

#validating the proportion of splitting
```{r}
prop.table(table(train_data$Churn))
prop.table(table(test_data$Churn))
```

#feature importance using Boruta
```{r}
boruta_output <- Boruta(Churn~., data = telcom,
                        doTrace=2)
print(boruta_output)
```


RANDOM FOREST
#random forest classification model

```{r}
forest1 <- randomForest(Churn~., data = train_data)
forest1
```


#feature importance in the random forest model
```{r}
randomForest::importance(forest1)
```

```{r}
forestprediction <- predict(forest1, data = test_data[,19])
summary(forestprediction)
```


#LOGISTIC REGRESSION

```{r}
Churnmodel1 <- glm( Churn~., data = train_data, family = binomial(link = "logit"))
```

```{r}
summary(Churnmodel1)
```

FEATURE SELECTION

```{r}
Churnmodel2 <- glm(Churn~Dependents+Partner+SeniorCitizen+tenure+MultipleLines+OnlineSecurity+OnlineBackup+InternetService+TechSupport+DeviceProtection+Contract+StreamingMovies+StreamingTV+PaperlessBilling+PaymentMethod+MonthlyCharges, data = train_data, family = binomial(link = "logit"))
```

```{r}
summary(Churnmodel2)
```

```{r}
Churnmodel3 <- glm(Churn~gender+Partner+SeniorCitizen+Dependents+tenure+MultipleLines+InternetService+OnlineBackup+OnlineSecurity+DeviceProtection+TechSupport+Contract+PaperlessBilling+PaymentMethod+MonthlyCharges, data = train_data, family = binomial(link = "logit"))

```

```{r}
summary(Churnmodel3) 
```


```{r}
Churnmodel4 <- glm(Churn~Partner+Dependents+SeniorCitizen+tenure+MultipleLines+InternetService+OnlineSecurity+OnlineBackup+DeviceProtection+TechSupport+Contract+PaperlessBilling+PaymentMethod+MonthlyCharges, data = train_data, family = binomial(link = "logit"))
```

```{r}
summary(Churnmodel4)
```

#Comparing AIC values of models
```{r}
Model_AIC <- data.frame(AIC(Churnmodel1),AIC(Churnmodel2),AIC(Churnmodel3),AIC(Churnmodel4))
```

```{r}
view(Model_AIC)
```

```{r}
anova(Churnmodel4, test = "Chisq")
```

```{r}
car::vif(Churnmodel4)
```

#using the candidate model to predict outcomes in the test data
```{r}
predicted <- predict(Churnmodel4, test_data, "link")
head(predicted)
```

#saving the result as predicted
```{r}
test_data$predicted <- predicted
```

#creating a new column to compare actual and predicted
```{r}
test_data$Churn_pred <- ifelse(predicted>0.5, "Yes", "No")
```

```{r}
view(test_data)
```

REPEATED CROSS VALIDATION
 
```{r}
library(e1071) 
```

#repeated cross validation
```{r}
logChurnModel <- train(Churn~Partner+Dependents+SeniorCitizen+tenure+MultipleLines+InternetService+OnlineSecurity+OnlineBackup+DeviceProtection+TechSupport+Contract+PaperlessBilling+PaymentMethod+MonthlyCharges, data = train_data, method = "glm", family = "binomial",
                          trControl = trainControl(method = "repeatedcv", number = 20, repeats = 10))
```

#output model from repeated cross validation
```{r}
logChurnModel
summary(logChurnModel)
```


#Prediction using trained model against test data
```{r}
logChurnPrediction <- predict(logChurnModel, test_data)
```

#getting a summary statistics of the prediction
```{r}
summary(logChurnPrediction)
```


#Getting detailed statistics of PREDICTION VS ACTUAL via confusion matrix
```{r}
logChurnConfMat <- confusionMatrix(logChurnPrediction, test_data[,"Churn"])
logChurnConfMat
```

```{r}

```

