---
title: "Telco customer Churn Prediction"
author: "Precious Ebu"
date: "11/20/2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# core tidyverse + modelling tools

library(tidyverse)
library(psych)
library(ggpubr)
library(ggplot2)
library(readr)
library(stringr)
library(lattice)
library(caret)
library(randomForest)
library(Boruta)
library(car)
library(pROC)
library(vip)
library(xgboost)
library(Matrix)

```

# ---- Load Data ----
```{r}
# Use a relative path if possible; update to your path

data_path <- "C:\\Users\\hp\\OneDrive\\Documents\\Data Science\\Datasets\\Telcom Dataset - Copy.csv"

telcom <- read_csv(data_path, na = c("", " ", "NA"))

```

```{r}
# ----- Quick Inspect -----
glimpse(telcom)
summary(telcom)
```

# ---- Quick inspection of categorical values ----
```{r}
# Quick unique value checks — helps find inconsistent labels

cat_cols <- c("gender","SeniorCitizen","Partner","Dependents","PhoneService","MultipleLines",
"InternetService","OnlineSecurity","OnlineBackup","DeviceProtection","TechSupport",
"StreamingTV","StreamingMovies","Contract","PaperlessBilling","PaymentMethod","Churn")

# Only print unique values for existing columns

cat_cols <- cat_cols[cat_cols %in% names(telcom)]

sapply(telcom[cat_cols], function(x) unique(x) %>% head(10))

```

# ----- Data Cleaning & Type conversions -----

```{r}
# Convert known categorical columns to factors (preserve original names like in your dataset)

for (col in intersect(cat_cols, names(telcom))) {
telcom[[col]] <- as.factor(telcom[[col]])
}

# Ensure numeric columns are numeric

telcom$tenure <- as.integer(telcom$tenure)
telcom$MonthlyCharges <- as.numeric(telcom$MonthlyCharges)

# Handle TotalCharges safely: only parse/convert if it's character like common telco CSVs

if ("TotalCharges" %in% names(telcom)) {
if (!is.numeric(telcom$TotalCharges)) {
telcom$TotalCharges <- parse_number(as.character(telcom$TotalCharges))
} else {
telcom$TotalCharges <- as.numeric(telcom$TotalCharges)
}
}

# Quick checks

glimpse(telcom)
colSums(is.na(telcom))

```


# ---- Basic cleaning (fix known inconsistencies) ----

```{r}
# Fix common inconsistent strings (case and phrase variations)

telcom <- telcom %>%
mutate(
InternetService = ifelse(is.na(InternetService), "No", as.character(InternetService)),
InternetService = str_replace_all(InternetService, regex("dsl", ignore_case = TRUE), "DSL"),
Contract = str_replace_all(as.character(Contract), regex("2 year|2 years", ignore_case = TRUE), "Two year"),
PaymentMethod = str_replace_all(as.character(PaymentMethod), "Electronic_check", "Electronic check"),
OnlineSecurity = str_replace_all(as.character(OnlineSecurity), regex("NO|No Internet Service", ignore_case = TRUE), "No"),
OnlineBackup = str_replace_all(as.character(OnlineBackup), regex("NO", ignore_case = TRUE), "No")
)

# Re-factor updated columns (only necessary ones)

telcom <- telcom %>%
mutate(
InternetService = as.factor(InternetService),
Contract = as.factor(Contract),
PaymentMethod = as.factor(PaymentMethod),
OnlineSecurity = as.factor(OnlineSecurity),
OnlineBackup = as.factor(OnlineBackup)
)

```

# --- Missing value handling (targeted and documented) ---
```{r}
# Examine missingness

colSums(is.na(telcom))
```

```{r}
# Replace remaining numeric NAs with median
num_vars <- telcom %>% select(where(is.numeric)) %>% names()
for (nv in num_vars) {
  med <- median(telcom[[nv]], na.rm = TRUE)
  telcom[[nv]][is.na(telcom[[nv]])] <- med
}
```


```{r}
# Replace factor NAs with mode
mode_replace <- function(x){
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x, ux)))]
}
fac_vars <- telcom %>% select(where(is.factor)) %>% names()
for (fv in fac_vars) {
  telcom[[fv]][is.na(telcom[[fv]])] <- mode_replace(telcom[[fv]])
}

colSums(is.na(telcom)) #confirm replacement of missing values
```


# --- Outlier handling ---
```{r}

# detect extreme tenure values and cap them at reasonable upper bound (99th percentile)

if ("tenure" %in% names(telcom)) {
q99 <- quantile(telcom$tenure, 0.99, na.rm = TRUE)
telcom$tenure <- ifelse(telcom$tenure > q99, round(q99), telcom$tenure)
}
boxplot(telcom$tenure, main = "Tenure after capping")

```

# ----- Feature Engineering -----

```{r}
# Compute TotalChargesCal; also create avg monthly charge and tenure buckets

telcom <- telcom %>%
mutate(
TotalChargesCal = tenure * MonthlyCharges,                 # same idea you had
AvgMonthlyCharge = ifelse(tenure > 0, TotalCharges / tenure, MonthlyCharges),
TenureGroup = cut(tenure,
breaks = c(-1, 6, 12, 24, 48, Inf),
labels = c("0-6","7-12","13-24","25-48","49+"),
right = TRUE)
) %>%
mutate(TenureGroup = factor(TenureGroup, levels = c("0-6","7-12","13-24","25-48","49+")))

```

# ----- EDA -----

```{r}
# Churn proportion

telcom %>% count(Churn) %>% mutate(pct = n / sum(n) * 100)

#,Plots: churn by tenure and monthly charges

p1 <- ggplot(telcom, aes(x = tenure, fill = Churn)) +
geom_histogram(binwidth = 1.5) + labs(title = "Churn by Tenure (months)")

p2 <- ggplot(telcom, aes(x = MonthlyCharges, fill = Churn)) +
geom_histogram(binwidth = 2) + labs(title = "Churn by Monthly Charges")

# Print figures (for Rmd)

print(p1)
print(p2)
```


```{r}
# Bar plots for selected categorical variables (loop)

vars <- c("Partner","Dependents","InternetService","Contract","PaymentMethod")
for (v in vars[vars %in% names(telcom)]) {
print(
ggplot(telcom, aes_string(x = v, fill = "Churn")) +
geom_bar(position = "stack") +
theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
labs(title = paste("Churn by", v))
)
}

```


```{r}
# Correlations (numeric)
num_cor <- cor(telcom %>% select(where(is.numeric)))
```


# --- Prepare data for modeling — stratified split ---

```{r}
# ----- Train/Test Split (stratified) -----
set.seed(1500)

# Use caret's createDataPartition to stratify by Churn (preserve class distribution)

if (!"Churn" %in% names(telcom)) stop("Churn column missing")

train_idx <- createDataPartition(telcom$Churn, p = 0.75, list = FALSE)
train_data <- telcom[train_idx, ]
test_data  <- telcom[-train_idx, ]

# Check proportions

prop.table(table(train_data$Churn))
prop.table(table(test_data$Churn))

```

# --- Feature selection exploratory (Boruta) ---

```{r}
# Run Boruta on train set to get a sense of important features 

set.seed(42)
boruta_output <- Boruta(Churn ~ ., data = train_data %>% select(-customerID), doTrace = 0)
print(boruta_output)
plot(boruta_output, las = 2, cex.axis = 0.6)

```

# --- Random Forest — training & evaluation ---
```{r}
# Ensure Churn factor levels have "No"/"Yes" in this order for caret's twoClassSummary

train_data$Churn <- factor(train_data$Churn, levels = c("No","Yes"))
test_data$Churn  <- factor(test_data$Churn, levels = c("No","Yes"))

# caret train control (repeated cv)

ctrl <- trainControl(method = "repeatedcv",
number = 5,
repeats = 3,
classProbs = TRUE,
summaryFunction = twoClassSummary,
savePredictions = "final",
verboseIter = FALSE)

# tune grid for random forest (mtry)

mtry_grid <- expand.grid(mtry = c(2, 4, 6, 8))

set.seed(42)
rf_model <- train(Churn ~ Partner + Dependents + SeniorCitizen + tenure + MultipleLines +
InternetService + OnlineSecurity + OnlineBackup + DeviceProtection +
TechSupport + Contract + StreamingMovies + StreamingTV + PaperlessBilling +
PaymentMethod + MonthlyCharges + TotalChargesCal + AvgMonthlyCharge + TenureGroup,
data = train_data,
method = "rf",
metric = "ROC",
tuneGrid = mtry_grid,
trControl = ctrl,
importance = TRUE,
ntree = 500)

rf_model

```


```{r}
# Evaluate RF on test set

rf_probs <- predict(rf_model, test_data, type = "prob")[, "Yes"]
rf_pred  <- factor(ifelse(rf_probs > 0.5, "Yes", "No"), levels = c("No","Yes"))

conf_rf <- confusionMatrix(rf_pred, test_data$Churn, positive = "Yes")
conf_rf

# ROC / AUC

roc_rf <- roc(response = test_data$Churn, predictor = rf_probs, levels = c("No","Yes"))
auc_rf <- auc(roc_rf)
auc_rf
plot(roc_rf, main = paste("Random Forest ROC — AUC:", round(auc_rf, 3)))

```

```{r}
# Variable importance (caret/vip)

vip::vip(rf_model$finalModel, num_features = 15) + ggtitle("RF - Variable Importance (finalModel)")

```

# ----- Logistic Regression (glm) -----

```{r}
# Fit a full logistic regression and use stepwise selection (AIC) to find a candidate model

glm_full <- glm(Churn ~ ., data = train_data %>% select(-customerID), family = "binomial")
glm_step <- step(glm_full, direction = "both", trace = FALSE)

summary(glm_step)

# Show selected formula

formula(glm_step)
```


```{r}
# Train the selected glm using caret with repeated CV for better generalization

set.seed(42)
log_ctrl <- trainControl(method = "repeatedcv",
number = 5,
repeats = 3,
classProbs = TRUE,
summaryFunction = twoClassSummary)

log_model <- train(formula(glm_step), data = train_data,
method = "glm",
family = "binomial",
metric = "ROC",
trControl = log_ctrl)

log_model

```

```{r}
# Evaluate logistic on test set

log_probs <- predict(log_model, test_data, type = "prob")[, "Yes"]
log_pred  <- factor(ifelse(log_probs > 0.5, "Yes", "No"), levels = c("No","Yes"))

conf_log <- confusionMatrix(log_pred, test_data$Churn, positive = "Yes")
conf_log

roc_log <- roc(response = test_data$Churn, predictor = log_probs, levels = c("No","Yes"))
auc_log <- auc(roc_log)
auc_log
plot(roc_log, main = paste("Logistic Regression ROC — AUC:", round(auc_log, 3)))

```

# ---- XGBoost Model----

```{r}
# --- Create consistent matrices for XGBoost ---

# 1. Build training matrix
train_matrix <- model.matrix(Churn ~ . - 1, data = train_data)

# 2. Build test matrix (raw)
test_matrix_raw <- model.matrix(Churn ~ . - 1, data = test_data)

# 3. Add missing columns to test
missing_cols <- setdiff(colnames(train_matrix), colnames(test_matrix_raw))
for (mc in missing_cols) {
  test_matrix_raw <- cbind(test_matrix_raw, rep(0, nrow(test_matrix_raw)))
  colnames(test_matrix_raw)[ncol(test_matrix_raw)] <- mc
}

# 4. Align test matrix EXACTLY to training
test_matrix <- test_matrix_raw[, colnames(train_matrix)]

# --- Labels ---
train_label <- ifelse(train_data$Churn == "Yes", 1, 0)
test_label  <- ifelse(test_data$Churn == "Yes", 1, 0)
```


```{r}
# --- Convert to DMatrix ---
dtrain <- xgb.DMatrix(data = train_matrix, label = train_label)
dtest  <- xgb.DMatrix(data = test_matrix,  label = test_label)
```


```{r}
# --- XGBoost Parameters ---
params <- list(
  booster = "gbtree",
  objective = "binary:logistic",
  eval_metric = "auc",
  eta = 0.05,
  max_depth = 6,
  subsample = 0.8,
  colsample_bytree = 0.8
)
```


```{r}
# --- Train XGBoost Model ---
set.seed(42)
xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 300,
  watchlist = list(train = dtrain, test = dtest),
  verbose = 0
)
```


```{r}
# --- Predictions ---
xgb_pred_prob <- predict(xgb_model, dtest)
xgb_pred_class <- ifelse(xgb_pred_prob > 0.5, "Yes", "No")
xgb_pred_class <- factor(xgb_pred_class, levels = c("No", "Yes"))
```


```{r}
# --- Confusion Matrix ---
xgb_conf_mat <- caret::confusionMatrix(
  xgb_pred_class,
  test_data$Churn,
  positive = "Yes"
)
xgb_conf_mat
```


```{r}
# --- ROC / AUC ---
xgb_roc <- roc(test_label, xgb_pred_prob)
auc_xgb <- auc(xgb_roc)
auc_xgb
plot(xgb_roc, main = paste("XGBoost ROC Curve — AUC:", round(auc_xgb, 3)))
```


```{r}
# --- XGBoost Feature Importance ---
xgb_imp <- xgb.importance(model = xgb_model, feature_names = colnames(train_matrix))
xgb.plot.importance(xgb_imp, top_n = 20)
```


```{r}
# Save AUC for comparison section
xgb_test_auc <- auc_xgb

```


# ---- XGBoost ROC Curve ---
```{r}
xgb_roc <- roc(response = test_data$Churn, predictor = xgb_pred_prob, levels = c("No","Yes"))

```

# --- Compare models and summary metrics ---
```{r}
# ---- Compare Models by Test AUC ----

model_results <- tibble(
  Model = c("Logistic Regression", "Random Forest", "XGBoost"),
  Test_AUC = c(auc_log, auc_rf, auc(xgb_roc))
)

model_results <- model_results %>% arrange(desc(Test_AUC))
model_results

```

# --- Final Model Selection ---
```{r}
best_model <- model_results$Model[1]

if (best_model == "Random Forest") {
  final_model <- rf_model
  final_probs <- rf_probs
} else if (best_model == "Logistic Regression") {
  final_model <- log_probs
  final_probs <- log_probs
} else {
  final_model <- xgb_model
  final_probs <- xgb_pred_prob
}

best_model

```

